---
output: github_document
editor_options: 
  chunk_output_type: console
bibliography: references.bib
link-citations: yes
linkcolor: Blue
urlcolor: Blue
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_chunk$set(fig.width=10, fig.height=4)
options(tidyverse.quiet = TRUE)
```
# epichannel

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/epichannel)](https://cran.r-project.org/package=epichannel)
<!-- badges: end -->

The goal of epichannel is to create classical endemic channel for Epidemiological Surveillance in Public Health.

## Installation

<!-- You can install the released version of epichannel from [CRAN](https://CRAN.R-project.org) with: install.packages("epichannel") -->

``` r
if(!require("devtools")) install.packages("devtools")
devtools::install_github("avallecam/epichannel")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(epichannel)
## basic example code
```

- First, _read_ surveillance data.

```{r,warning=FALSE,message=FALSE}
library(tidyverse)

# disease dataset
dengv <-
  readr::read_csv("https://dengueforecasting.noaa.gov/Training/Iquitos_Training_Data.csv") %>%
  mutate(year = lubridate::year(week_start_date),
         epiweek = lubridate::epiweek(week_start_date)) %>%
  mutate(adm="iquitos") %>%
  # cases per season - replace wiht a dummy year
  mutate(year = str_replace(season,"(.+)/(.+)","\\1") %>% as.double())

dengv %>% glimpse()

# population dataset
popdb <-
  readr::read_csv("https://dengueforecasting.noaa.gov/PopulationData/Iquitos_Population_Data.csv") %>%
  janitor::clean_names() %>%
  mutate(adm="iquitos")

popdb %>% glimpse()
```

```{r,echo=FALSE,eval=FALSE}
# dengv %>% count(year,season,lag_year)
# dengv %>%
#   ggplot(aes(x = week_start_date,y = total_cases)) +
#   geom_col()
# popdb %>% count(year)
# dengv %>% count(year)
# dengv %>% left_join(popdb)
```

- Second, _adapt_ both datasets

```{r}
epi_adapted <-
  epi_adapt_timeserie(db_disease = dengv,
                      db_population = popdb,
                      var_admx = adm,
                      var_year = year, # must be a common variable name between datasets
                      var_week = season_week,
                      # var_year = year, 
                      # var_week = epiweek,
                      var_event_count = total_cases,
                      var_population = estimated_population)
```

- Third, _filter_ by year to discriminate between the historical and current data 

```{r}
disease_now <- epi_adapted %>%
  filter(var_year==max(var_year))

disease_pre <- epi_adapted %>%
  filter(var_year!=max(var_year))
```

- Fourth, _create_ the __endemic channel__

  + here you can choose between three available methods [@bortman1999elaboracion]:
    
    + `"gmean_1sd"` is geometric mean with 1 standard deviation (default). 
    + `"gmean_2sd"` is geometric mean with 2 sd. 
    + `"gmean_ci"` is geometric mean with 95 percent confidence intervals.

```{r}
disease_channel <-
  epi_create_channel(time_serie = disease_pre,
                     disease_name = "dengv",
                     method = "gmean_1sd")

disease_channel

```

- Finally, _join_ datasets and _plot_ it with ggplot

```{r}
epi_join_channel(disease_channel = disease_channel,
                 disease_now = disease_now) %>%
  # ggplot
  epi_plot_channel() +
  labs(title = "DENGV Endemic Channel. Iquitos, Peru 2008/2009",
       caption = "Source: https://dengueforecasting.noaa.gov/",
       # x = "epiweeks",
       x = "Seasonal week",
       y = "Number of cases") +
  theme_bw()
```

## Contribute

Feel free to fill an issue with a [`reprex`](https://reprex.tidyverse.org/) or implement new methods through a pull request. Here are some alternatives:

- @orellano2011nuevo
- @Hernández_Arboleda_Arce_Benavides_Tejada_Ramírez_Cubides_2016
- @Bowman2016

## More

For a more advanced approach into surveillance algorithms refer to the `surveillance` R package and associated publications:

- @hohle2007tt
- @monitoring2016
- @modeling2017
- @germany2016

## References


